Tropical Mathematics and the Lambda Calculus

### 2023-01-22 appIII.tex
commit

\subsubsection{The language $\BSTLC$}
In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with affine simple types and an exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a fragment of $\mathrm{Fuzz}$ \cite{Reed2010}. 
We describe the rules of this fragment in Fig.~\ref{fig:fuzz}.
The terms are generated by the grammar:

{
\begin{minipage}{\textwidth}
\begin{align*}
t,u:= x\mid \lambda x.t\mid tu \mid !t \mid \mathsf{let}\  !x=t \ \mathsf{in}\  u\mid (t,u)\mid  \mathsf{let}\  (x,y)=t \ \mathsf{in} \  u 
\end{align*}\end{minipage}}\medskip\\
For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed2010,Gaboardi2017}.  
Simple types are generated by the grammar below:

{
\begin{minipage}{\textwidth}
\begin{align*}
A,B:= X\mid  !_{n}A  \mid A\multimap B \mid A\otimes B
\end{align*}\end{minipage}}\medskip\\
Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x\in_{n}A$.
We define the following operation $\Gamma+\Delta$  as follows:

{
\begin{minipage}{\textwidth}
\begin{align*}
() + () & =() \\
(\Gamma, x\in_{m} A)+( \Delta, x\in_{n} A) & =  (\Gamma+\Delta), x\in_{m+n}A \\
(\Gamma, x\in_{m}A)+\Delta & =(\Gamma+\Delta), x\in_{n}A \qquad (x\notin \Delta) \\
\Gamma+ (\Delta, x\in_{m}A) &= (\Gamma+\Delta), x\in_{n} A \qquad (x\notin \Gamma)
\end{align*}\end{minipage}}\medskip\\
Moreover, we let $m\Gamma$ be the context made all judgmenets $x\in_{mn}A$, where $(x\in_{n}A)\in \Gamma$.

Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
$!_{2}(A\multimap A)\multimap (A\multimap A)$.

@@

  {
\AXC{$x\in_{n} A\in \Gamma$}
\UIC{$\Gamma\vdash x\in A$}
\DP}

@@

\begin{tabular}{c c }
  {
\AXC{$\Gamma, x\in_{1} A\vdash t\in B$}
\UIC{$\Gamma\vdash \lambda x.t\in A\multimap B$}
\DP}

@@

  {
\AXC{$\Gamma \vdash t\in A\multimap B$}
\AXC{$\Gamma\vdash u\in A$}
\BIC{$\Gamma \vdash tu\in B$}
\DP}
\\

@@

  {
\AXC{$\Gamma \vdash t\in A$}
\AXC{$\Delta \vdash u\in B$}
\BIC{$\Gamma+\Delta \vdash ( t,u) \in A\otimes B$}
\DP}

@@

  {
\AXC{$\Gamma\vdash t\in A \otimes B$}
\AXC{$\Delta,x\in_{n}A, y\in_{n} B\vdash u:C$}
\BIC{$n\Gamma+\Delta \vdash \mathsf{let}\  (x,y)=t \ \mathsf{in}  \ u\in C$}
\DP}
\\
&

  {
\AXC{$\Gamma \vdash t\in A$}
\UIC{$s\Gamma \vdash !t \in \ !_{n}A$}
\DP}

  {
\AXC{$\Gamma\vdash t\in !_{n}A$}
\AXC{$\Delta,x\in_{mn}A\vdash u:C$}
\BIC{$m\Gamma+\Delta \vdash \mathsf{let}\  !x=t \ \mathsf{in} \ u\in C$}
\DP}

\end{tabular}
\end{center}
\end{minipage}
}
\caption{Typing rules for $\BSTLC$.}
\label{fig:fuzz}
\end{figure}

\subsubsection{$(\LREL, !_{n})$ is a model of Bounded Linear Logic}

@@

>>>>

\subsubsection{The language $\BSTLC$}
In Section 4 we considered a variant $\BSTLC$ of $\STLC$ with linear simple types and a graded exponential $!_{n}A$, for all $n\in \BB N$, corresponding to a somehow simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
%The terms are generated by the grammar:
%
%{
%\begin{minipage}{\textwidth}
%\begin{align*}
%M:= x\mid \lambda x.M\mid MM \mid !M \mid \mathsf{let}\  !x=M \ \mathsf{in}\  M%\mid (t,u)\mid  \mathsf{let}\  (x,y)=t \ \mathsf{in} \  u 
%\end{align*}\end{minipage}}\medskip\\
%For the purposes of this article we limited ourselves to a minimal fragment of this language. For a more practical language see \cite{Reed2010,Gaboardi2017}.

The types are generated by the grammar below:

{
\begin{minipage}{\textwidth}
\begin{align*}
A:= X \ \mid  \ !_{n}A \multimap A
\end{align*}\end{minipage}}\medskip\\
Type judgements are of the form $\Gamma \vdash t:A$, where a context $\Gamma$ is a list of declarations of the form $x :_{n}A$, with $n\in \BB N$.
We define the following operation $\Gamma+\Delta$  as follows:

{
\begin{minipage}{\textwidth}
\begin{align*}
() + () & =() \\
(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) & =  (\Gamma+\Delta), x:_{m+n}A \\
(\Gamma, x:_{n}A)+\Delta & =(\Gamma+\Delta), x:_{n}A \qquad (x\notin \Delta) \\
\Gamma+ (\Delta, x:_{n}A) &= (\Gamma+\Delta), x:_{n} A \qquad (x\notin \Gamma)
\end{align*}\end{minipage}}\medskip\\
Moreover, we let $m\Gamma$ be the context made all judgements $x:_{mn}A$, where $(x:_{n}A): \Gamma$.

The typing rules of $\BSTLC$ are illustrated in Fig.~\ref{fig:fuzz}.

%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
%$!_{2}(A\multimap A)\multimap (A\multimap A)$.

@@

  {
\AXC{}
\UIC{$ x:_{1}A\vdash x: A$}
\DP}

@@

\begin{tabular}{c c }
\AXC{$\Gamma \vdash M:A$}
\UIC{$\Gamma, x:_{0}B \vdash M:A$}
\DP

@@

\AXC{$\Gamma, x:_{n}B, y_{m}:B\vdash M:A$}
\UIC{$\Gamma, x:_{n+m}B\vdash M[x/y]:A$}
\DP

@@

  {
\AXC{$\Gamma, x:_{n} A\vdash M: B$}
\UIC{$\Gamma\vdash \lambda x.M: !_{n}A\multimap B$}
\DP}

@@

  {
\AXC{$\Gamma \vdash M: A\multimap B$}
\AXC{$\Delta\vdash N: A$}
\BIC{$\Gamma +\Delta\vdash MN: B$}
\DP}
\\

%&

%\\
%
%  {
%\AXC{$\Gamma \vdash t\in A$}
%\AXC{$\Delta \vdash u\in B$}
%\BIC{$\Gamma+\Delta \vdash ( t,u) \in A\otimes B$}
%\DP}
%
%& 
%
%
%  {
%\AXC{$\Gamma\vdash t\in A \otimes B$}
%\AXC{$\Delta,x\in_{n}A, y\in_{n} B\vdash u:C$}
%\BIC{$n\Gamma+\Delta \vdash \mathsf{let}\  (x,y)=t \ \mathsf{in}  \ u\in C$}
%\DP}
%\\
%&

%\\
%
%
%
%
%
%& 
%
%  {
%\AXC{$\Gamma\vdash M: !_{n}A$}
%\AXC{$\Delta,x\in_{mn}A\vdash u:C$}
%\BIC{$m\Gamma+\Delta \vdash \mathsf{let}\  !x=M \ \mathsf{in} \ u :  C$}
%\DP}

\end{tabular}

  {
\AXC{$\Gamma \vdash M: A$}
\UIC{$n\Gamma \vdash M : \ !_{n}A$}
\DP}

\end{center}
\end{minipage}
}
\caption{Typing rules for $\BSTLC$.}
\label{fig:fuzz}
\end{figure}

\subsubsection{$(\LREL, !_{n})$ is a model of $\BSTLC$}

@@

The following definition provides an interpretation of $\BSTLC$ in any symmetric monoidal closed category with a $\BB N$-graded linear exponential comonad.

\begin{definition}[interpretation of $\BSTLC$]
Let $\C C$ be a symmetric monoidal closed category and $(D, w,c,\epsilon,\delta)$ be a $\BB N$-graded linear exponential comonad. 
Let $\model{X}$ be fixed objects of $\C C $, one for each ground type $X$ of $\BSTLC$.

One lifts the interpretation to types as 
$\model{!_{n}A\multimap B}=D(n)(\model{A})\multimap \model B$. Moreover, one extends the interpretation to contexts via $\model{x:_{n}A}:= D(n)(\model A)$ and 
$\model{\{x_{1}:_{n_{1}}A_{1},\dots, x_{k}:_{n_{k}}A_{k}\}}=
\bigotimes_{i=1}^{k}\model{x_{i}:_{n_{i}}A_{i}}$.

Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in \C C(\model\Gamma, \model A)$ of $\Gamma \vdash M:A$ by induction as follows:
 \begin{itemize}
\item $\model{x\in_{1}A\vdash x\in A}=\epsilon_{A}$;
\item $\model{\Gamma+x:_{0}B\vdash M:A} =
\model{\Gamma \vdash M:A}
\circ (\model{\Gamma}\otimes  w_{\model B})
$;
\item $\model{\Gamma, x:_{m+n}B \vdash M[x/y]:A}=
\model{\Gamma, x:_{m}B, y:_{m}B \vdash M:A}
\circ
(\model\Gamma \otimes c_{m,n,\model B})
$;
\item $\model{\Gamma \vdash \lambda x.M: !_{n}A\multimap B}=
\Lambda (\model{\Gamma, x:_{n}A \vdash M:B})
$, where $\Lambda$ is the isomorphism $ \C C(\model \Gamma \otimes D(n)(\model A), \model B) \to \C C(\model \Gamma, D(n)(\model A)\multimap \model B)$;

\item $\model{\Gamma+\Delta \vdash MN:B}=
\mathsf{ev}\circ \big( \model{\Gamma \vdash M:A\multimap B}
\otimes
\model{\Delta \vdash N:A}\big)$, where $\mathsf{ev}\in \C C((\model{A}\multimap \model B)\otimes \model A, \model B)$ is the evaluation morphism of $\C C$;

\item $\model{n\Gamma\vdash M:!_{n}A}=
 !_{n}(\model{\Gamma \vdash M:A})\circ \big(\delta_{n,m_{1},\model{A_{1}}}\otimes \dots \otimes 
\delta_{n,m_{k},\model{A_{k}}}\big)
$, where $\Gamma=\{x_{1}:_{m_{1}}A_{1},\dots, x_{k}:_{m_{k}}A_{k}\}$.
%\item $\model{m\Gamma+\Delta \vdash \mathsf{let}\ !x=M\ \mathsf{in} \ u:C}=
%\model{\Delta,x:_{mn}A\vdash u:C}
%\circ
%!_{m}(\model{\Gamma \vdash M:!_{n}A})
%$.

\end{itemize}
\end{definition}

Let us now show that bounded multisets defined a $\BB N$-graded linear exponential comonad over $\LREL$.

### 2023-01-22 appIV.tex
commit

One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{ER}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).

>>>>

One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{difflambda}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).

### 2023-01-22 section4.tex
commit

A tropical polynomial is a piece-wise linear function of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

%\begin{remark}
Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
% \end{remark}
 \begin{remark}
 Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
% By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not discrete.
% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{remark}

>>>>

A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
are illustrated in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes). For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

%\begin{remark}
Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
% \end{remark}
 \begin{remark}
 Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
% By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not discrete.
% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{remark}

### 2023-01-22 section6.tex
commit

Beyond limits of Cauchy sequences, another important example of colimit is the following:
\begin{definition}
Let $X$ be a $\Lawv$-category, $x\in X$ and $\epsilon \in \Lawv$. The \emph{tensor of $x$ and $\epsilon$}, if it exists, is the colimit $\epsilon \otimes x:= \colim( [\epsilon],\Delta x)$, where
$[\epsilon]: \B 1\pfun \B 1$ is the constantly $\epsilon$ distributor
and $\Delta x:\B 1\to X$ is the constant functor. 
\end{definition}

>>>>

Beyond limits of Cauchy sequences, another important example of colimit is the following:
\begin{definition}
Let $X$ be a $\Lawv$-category, $x\in X$ and $\epsilon \in \Lawv$. The \emph{tensor of $x$ and $\epsilon$}, if it exists, is the colimit $\epsilon \otimes x:= \colim( [\epsilon],\Delta x)$, where
$[\epsilon]: \{\star\}\pfun \{\star\}$ is the constantly $\epsilon$ distributor
and $\Delta x:\{\star\}\to X$ is the constant functor. 
\end{definition}
